---
title: "LHS Case Study: Replicated Design Comparison"
---

## Setup

```{r}
renv::load("/Users/max/r_package_development/bayes2stage")
library(bayes2stage)

SCRIPT_DIR <- "."
source("lhs_config.R")
```


## Load Data

```{r}
df <- load_lhs_data()

N_SUBJECTS <- length(unique(df$id))
N_SAMPLED <- as.integer(SAMPLING_FRACTION * N_SUBJECTS)
cli::cli_alert_info("N subjects: {N_SUBJECTS}, N sampled (stage 2): {N_SAMPLED}")
```


## lme4 Reference

```{r}
lme4_no_cov <- lme4::lmer(y ~ x + t + x:t + (1 + t | id), data = df)

lme4_schildcrout <- lme4::lmer(
    y ~ x + t + x:t +
        packyear + f31cigs + bmi0 + female + age + site +
        bmi_change + t:age + t:female + t:site +
        (1 + t | id),
    data = df
)

summary(lme4_no_cov)
summary(lme4_schildcrout)
```


## Replicated Design Comparison

### Load Results

Results are pre-computed by `run_rep.R` via `gnu_parallel.sh`.
Each replicate writes to `results/rep_{id}.parquet`.

```{r}
#| label: load-results

result_files <- fs::dir_ls("results", glob = "*.parquet")
cli::cli_alert_info("Found {length(result_files)} result files")

all_results <- purrr::map(result_files, arrow::read_parquet) |>
    dplyr::bind_rows()

# Report any fits that failed
if ("error" %in% names(all_results)) {
    fit_errors <- dplyr::filter(all_results, !is.na(error))
    if (nrow(fit_errors) > 0L) {
        cli::cli_alert_warning("{nrow(fit_errors)} fits failed:")
        print(dplyr::count(fit_errors, model_spec, design, method, name = "n_failed"))
    }
    all_results <- dplyr::filter(all_results, is.na(error)) |>
        dplyr::select(-error)
}

# Split results for downstream analysis
stan_results <- dplyr::filter(all_results, method == "stan")
full_results <- dplyr::filter(stan_results, design == "full")
rep_results <- dplyr::filter(stan_results, design != "full")

N_REPS_ACTUAL <- length(unique(rep_results$replicate))
cli::cli_alert_success(
    "Loaded {nrow(all_results)} rows: {N_REPS_ACTUAL} replicates + full-data fits"
)
```


### Aggregate Results

```{r}
#| label: aggregate

agg <- rep_results |>
    dplyr::mutate(ci_width = q97_5 - q2_5) |>
    dplyr::summarize(
        mean_of_means = mean(mean),
        sd_of_means = sd(mean),
        mean_ci_width = mean(ci_width),
        n_reps = dplyr::n(),
        .by = c(model_spec, design, parameter)
    )

full_ref <- full_results |>
    dplyr::mutate(full_ci_width = q97_5 - q2_5) |>
    dplyr::select(model_spec, parameter, full_mean = mean, full_ci_width)

agg <- dplyr::left_join(agg, full_ref, by = c("model_spec", "parameter"))
agg$bias <- agg$mean_of_means - agg$full_mean
```


## Results

```{r}
library(ggplot2)
library(gt)
theme_set(cowplot::theme_cowplot(font_size = 12))

fs::dir_create("figures")
```


### Summary Table

```{r}
#| label: tbl-summary

agg |>
    dplyr::filter(parameter %in% KEY_PARAMS) |>
    dplyr::mutate(
        design_label = DESIGN_LABELS[design],
        param_label = PARAM_LABELS[parameter],
        spec_label = SPEC_LABELS[model_spec]
    ) |>
    dplyr::select(
        spec_label, design_label, param_label,
        mean_of_means, sd_of_means, mean_ci_width, full_mean, bias
    ) |>
    gt(groupname_col = "spec_label") |>
    fmt_number(
        columns = c(mean_of_means, sd_of_means, mean_ci_width, full_mean, bias),
        decimals = 4
    ) |>
    cols_label(
        design_label = "Design",
        param_label = "Parameter",
        mean_of_means = "Mean(est)",
        sd_of_means = "SD(est)",
        mean_ci_width = "Mean(CI width)",
        full_mean = "Full data",
        bias = "Bias"
    )
```


### Forest Plot

```{r}
#| label: fig-forest
#| fig-width: 14
#| fig-height: 14

DESIGN_ORDER <- rev(c(
    "Full data", "SRS", "SRS (no imp)",
    "ODS (intercept)", "ODS (slope)",
    "BDS (intercept)", "BDS (slope)"
))

full_for_plot <- full_results |>
    dplyr::filter(parameter %in% KEY_PARAMS) |>
    dplyr::transmute(
        model_spec, parameter,
        design = "full",
        mean_of_means = mean,
        sd_of_means = 0
    )

forest_df <- agg |>
    dplyr::filter(parameter %in% KEY_PARAMS) |>
    dplyr::select(model_spec, parameter, design, mean_of_means, sd_of_means) |>
    dplyr::bind_rows(full_for_plot) |>
    dplyr::mutate(
        design_label = factor(DESIGN_LABELS[design], levels = DESIGN_ORDER),
        param_label = PARAM_LABELS[parameter],
        spec_label = SPEC_LABELS[model_spec]
    )

ref_lines <- full_results |>
    dplyr::filter(parameter %in% KEY_PARAMS) |>
    dplyr::mutate(
        param_label = PARAM_LABELS[parameter],
        spec_label = SPEC_LABELS[model_spec]
    )

p_forest <- forest_df |>
    ggplot(aes(x = mean_of_means, y = design_label)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
    geom_vline(
        data = ref_lines, aes(xintercept = mean),
        linetype = "dashed", color = "grey40"
    ) +
    geom_pointrange(aes(
        xmin = mean_of_means - sd_of_means,
        xmax = mean_of_means + sd_of_means
    )) +
    facet_grid(spec_label ~ param_label, scales = "free") +
    labs(
        title = glue::glue("Posterior mean estimates across {N_REPS_ACTUAL} replicated subsamples"),
        subtitle = "25% sampling fraction; dashed line = full-cohort estimate",
        x = glue::glue("Mean estimate \u00b1 SD across replicates"),
        y = NULL
    )
ggsave("figures/forest_replicates.png", p_forest,
    device = ragg::agg_png, bg = "white", dpi = 500,
    width = 14, height = 14, units = "in")
```


### Publication Forest Plot

```{r}
#| label: fig-forest-pub
#| fig-width: 12
#| fig-height: 5

agg_forest <- rep_results |>
    dplyr::summarize(
        est = mean(mean),
        se = mean((q97_5 - q2_5) / (2 * qnorm(0.975))),
        lo = mean(q2_5),
        hi = mean(q97_5),
        .by = c(model_spec, design, parameter)
    )

make_forest_data <- function(param, spec, agg_data, full_data) {
    f <- dplyr::filter(full_data, parameter == param, model_spec == spec)
    f_se <- (f$q97_5 - f$q2_5) / (2 * qnorm(0.975))

    get_row <- function(des) {
        row <- dplyr::filter(
            agg_data, design == des, parameter == param, model_spec == spec
        )
        if (nrow(row) == 0L) {
            return(list(est = NA_real_, se = NA_real_, lo = NA_real_, hi = NA_real_))
        }
        list(est = row$est, se = row$se, lo = row$lo, hi = row$hi)
    }

    d <- purrr::map(
        c("srs", "ods_intercept", "ods_slope", "bds_intercept", "bds_slope", "srs_no_imp"),
        get_row
    ) |> purrr::set_names(c("srs", "ods_i", "ods_s", "bds_i", "bds_s", "srs_ni"))

    se_vec <- c(f_se, NA, NA,
                d$srs$se, d$ods_i$se, d$ods_s$se,
                d$bds_i$se, d$bds_s$se,
                NA, NA, d$srs_ni$se)
    re_vec <- ifelse(is.na(se_vec), NA_real_, (f_se / se_vec)^2)

    data.frame(
        label = c(
            "Full cohort", "",
            "Bayesian imputation",
            "Random sampling (SRS)",
            "ODS (intercept)", "ODS (slope)",
            "BDS (intercept)", "BDS (slope)",
            "",
            "Complete-case (no imputation)",
            "Random sampling (SRS)"
        ),
        est = c(f$mean, NA, NA,
                d$srs$est, d$ods_i$est, d$ods_s$est,
                d$bds_i$est, d$bds_s$est,
                NA, NA, d$srs_ni$est),
        lo  = c(f$q2_5, NA, NA,
                d$srs$lo, d$ods_i$lo, d$ods_s$lo,
                d$bds_i$lo, d$bds_s$lo,
                NA, NA, d$srs_ni$lo),
        hi  = c(f$q97_5, NA, NA,
                d$srs$hi, d$ods_i$hi, d$ods_s$hi,
                d$bds_i$hi, d$bds_s$hi,
                NA, NA, d$srs_ni$hi),
        se  = se_vec,
        re  = re_vec
    )
}

combos <- tidyr::expand_grid(param = KEY_PARAMS, spec = names(MODEL_SPECS))

pub_plots <- purrr::map2(combos$param, combos$spec, \(param, spec) {
    fd <- make_forest_data(param, spec, agg_forest, full_results)
    fmt <- if (param == "beta_x_t_interaction") {
        list(ci = "%.4f (%.4f, %.4f)", se = "%.4f")
    } else {
        list(ci = "%5.2f (%5.2f, %4.2f)", se = "%4.2f")
    }
    zero_in_range <- min(fd$lo, na.rm = TRUE) <= 0 && max(fd$hi, na.rm = TRUE) >= 0
    p <- plot_forest(
        fd, ref_line = fd$est[1L],
        xlab = PARAM_LABELS[param],
        ci_format = fmt$ci, se_format = fmt$se,
        extra_col = list(col = "re", header = "RE", format = "%4.2f"),
        font_family = ""
    )
    if (zero_in_range) {
        p <- p + geom_vline(xintercept = 0, linetype = "dashed", color = "grey70")
    }
    p +
        ggtitle(
            glue::glue("{PARAM_LABELS[param]}: {SPEC_LABELS[spec]} model"),
            subtitle = glue::glue(
                "Mean posterior estimates averaged across {N_REPS_ACTUAL} replicates"
            )
        ) +
        theme(
            plot.title = element_text(size = 13, face = "bold"),
            plot.title.position = "plot",
            plot.subtitle = element_text(size = 10)
        )
})

purrr::walk2(
    pub_plots,
    glue::glue("figures/forest_{combos$spec}_{combos$param}.png"),
    \(p, path) ggsave(path, p,
        device = ragg::agg_png, bg = "white", dpi = 500,
        width = 12, height = 5, units = "in")
)

```


### Relative Efficiency

```{r}
#| label: fig-re
#| fig-width: 14
#| fig-height: 10

RE_DESIGN_ORDER <- rev(c(
    "SRS",
    "ODS (intercept)", "BDS (intercept)",
    "ODS (slope)", "BDS (slope)"
))

re_df <- agg |>
    dplyr::filter(
        parameter %in% KEY_PARAMS,
        !design %in% c("full", "srs_no_imp")
    ) |>
    dplyr::mutate(
        relative_efficiency = full_ci_width^2 / mean_ci_width^2,
        design_label = factor(DESIGN_LABELS[design], levels = RE_DESIGN_ORDER),
        param_label = PARAM_LABELS[parameter],
        spec_label = SPEC_LABELS[model_spec]
    )

p_re <- re_df |>
    ggplot(aes(x = relative_efficiency, y = design_label)) +
    geom_col() +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey30") +
    facet_grid(spec_label ~ param_label, scales = "free_x") +
    labs(
        title = glue::glue(
            "Relative efficiency vs. full cohort ({N_REPS_ACTUAL} replicates)"
        ),
        subtitle = "Values near 1 = minimal information loss; dashed line = full-cohort reference",
        x = "Relative efficiency (full CI width\u00b2 / design CI width\u00b2)",
        y = NULL
    )
ggsave("figures/relative_efficiency.png", p_re,
    device = ragg::agg_png, bg = "white", dpi = 500,
    width = 14, height = 10, units = "in")
```


### Bias

```{r}
#| label: fig-bias
#| fig-width: 14
#| fig-height: 10

BIAS_ORDER <- rev(c(
    "SRS", "SRS (no imp)",
    "ODS (intercept)", "ODS (slope)",
    "BDS (intercept)", "BDS (slope)"
))

bias_df <- agg |>
    dplyr::filter(parameter %in% KEY_PARAMS, design != "full") |>
    dplyr::mutate(
        design_label = factor(DESIGN_LABELS[design], levels = BIAS_ORDER),
        param_label = PARAM_LABELS[parameter],
        spec_label = SPEC_LABELS[model_spec]
    )

p_bias <- bias_df |>
    ggplot(aes(x = bias, y = design_label)) +
    geom_col() +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey30") +
    facet_grid(spec_label ~ param_label, scales = "free_x") +
    labs(
        title = glue::glue(
            "Bias relative to full-cohort estimate ({N_REPS_ACTUAL} replicates)"
        ),
        subtitle = "25% sampling fraction; dashed line = zero bias",
        x = "Bias (mean design estimate \u2212 full-cohort estimate)",
        y = NULL
    )
ggsave("figures/bias.png", p_bias,
    device = ragg::agg_png, bg = "white", dpi = 500,
    width = 14, height = 10, units = "in")
```


### ODS Slope Bias

ODS slope sampling selects subjects with extreme outcome trajectories, creating
informative missingness that the imputation model does not fully correct. Both
the Bayesian (Stan) and frequentist (ACML) estimators show similar bias for
$\beta_x$, confirming this is a property of the *design*, not the estimation
method.

```{r}
#| label: fig-ods-slope-bias
#| fig-width: 12
#| fig-height: 8

METHOD_LABELS <- c(stan = "Bayesian imputation (Stan)", acml = "ACML")

ods_slope_df <- all_results |>
    dplyr::filter(
        design %in% c("ods_slope", "full"),
        parameter == "beta_x"
    ) |>
    dplyr::mutate(
        method_label = METHOD_LABELS[method],
        spec_label = SPEC_LABELS[model_spec]
    )

full_ref_bx <- dplyr::filter(ods_slope_df, design == "full") |>
    dplyr::distinct(spec_label, mean)

ods_slope_rep <- dplyr::filter(ods_slope_df, design == "ods_slope")

p_ods_bias <- ods_slope_rep |>
    ggplot(aes(x = mean, color = method_label)) +
    geom_density(fill = NA, linewidth = 0.7) +
    geom_vline(
        data = full_ref_bx, aes(xintercept = mean),
        linetype = "dashed", linewidth = 0.6
    ) +
    facet_wrap(~ spec_label, scales = "free_x") +
    scale_color_manual(values = c(
        "Bayesian imputation (Stan)" = "#3574a7",
        "ACML" = "#e8833a"
    )) +
    labs(
        title = glue::glue(
            "ODS slope bias for \u03b2x: both methods agree ({N_REPS_ACTUAL} replicates)"
        ),
        subtitle = "Dashed line = full cohort",
        x = "Estimate of \u03b2x",
        y = "Density",
        color = "Method"
    ) +
    theme(legend.position = "bottom")
ggsave("figures/ods_slope_bias.png", p_ods_bias,
    device = ragg::agg_png, bg = "white", dpi = 500,
    width = 12, height = 8, units = "in")
```


## Diagnostics

```{r}
div_issues <- rep_results |>
    dplyr::distinct(model_spec, design, replicate, divergent_transitions) |>
    dplyr::filter(divergent_transitions > 0L)

if (nrow(div_issues) > 0L) {
    cli::cli_alert_warning("{nrow(div_issues)} fits had divergent transitions")
    print(dplyr::count(div_issues, model_spec, design, name = "n_fits_with_divs"))
} else {
    cli::cli_alert_success("No divergent transitions in any fit.")
}
```

```{r}
rhat_issues <- rep_results |>
    dplyr::filter(rhat > 1.01)

if (nrow(rhat_issues) > 0L) {
    cli::cli_alert_warning("{nrow(rhat_issues)} parameter-fit combos with Rhat > 1.01")
    print(dplyr::count(rhat_issues, model_spec, design, parameter, name = "n_reps"))
} else {
    cli::cli_alert_success("All Rhat <= 1.01")
}
```
