---
title: "Clean Data"
---

```{r}
library(tidyverse)
library(data.table)
```

```{r}
# Load in raw LHS data from Jonathan
load("lhsRohde.Rdata")
df <- lhsRohde
```

```{r}
# Check the overall data structure
Hmisc::describe(df)
```

## SNP columns

```{r}
# Identify the SNP columns -- start with "rs"
snps <- select(df, dplyr::starts_with("rs"))
```

```{r}
print(ncol(snps))
```

```{r}
if (anyNA(snps)) {
    cli::cli_abort("Some subjects have missing SNP values.")
}
```

# Data Cleaning

```{r}
# Threshold SNP dosages to 0/1
snps <- mutate(snps, across(everything(), \(x) as.integer(x >= 0.5)))

# Mean of SNPs within each subject
df$x_mean <- rowMeans(snps)

# Sum of SNPs within each subject
df$x_count <- rowSums(snps) |> as.integer()
```

```{r}
df$fev1 <- df$fev
df$t <- df$visit
df$fev_fvc_ratio <- df$fev / df$fvc

# BMI change from baseline (time-varying)
df$bmi_change <- df$bmi - df$bmi0

# Covariates
df$age <- df$age / 10
df$packyear <- df$packyear / 10
df$f31cigs <- df$f31cigs / 10

# Rename IDS to be sequential integers
df$id <- as.integer(as.factor(df$id))

df$rs177852 <- as.integer(df$rs177852 >= 0.5)

df <-
  df |>
  select(id,
         x_mean,
         x_count,
         rs177852,
         fev1,
         t,
         fevpct,
         fvc,
         fev_fvc_ratio,
         bmi_change,
         age,
         packyear,
         f31cigs,
         female,
         bmi0,
         site)
```

```{r}
arrow::write_parquet(df,
                     sink = "lhs_cleaned.parquet")
```

## List of variables in output data

- **id**: Subject identifier (sequential integers, 1 to N)
- **x_mean**: Polygenic risk score — unweighted mean of 43 hard-called SNPs (proportion, k/43)
- **x_count**: Polygenic risk score — sum of 43 hard-called SNPs (integer count, 0–43)
- **rs177852**: Binary indicator for presence of >= 1 T allele at rs177852 (*TTC6* gene)
- **fev1**: FEV1 in liters (forced expiratory volume in 1 second)
- **t**: Visit number (1–5, corresponding to annual visits)
- **fevpct**: FEV1 percent predicted (longitudinal, one per visit)
- **fvc**: Forced vital capacity in liters (longitudinal, one per visit)
- **fev_fvc_ratio**: FEV1/FVC ratio (computed as `fev / fvc`, longitudinal)
- **bmi_change**: BMI change from baseline (time-varying, `bmi - bmi0` at each visit)
- **age**: Age at baseline in decades (original years / 10)
- **packyear**: Cumulative pack-years of smoking in tens (original pack-years / 10)
- **f31cigs**: Cigarettes smoked per day in the year before enrollment, in tens (original count / 10)
- **female**: Sex indicator (1 = female, 0 = male)
- **bmi0**: Baseline body mass index (kg/m^2^, unscaled)
- **site**: Clinical site indicator (1–10)

## Correlation matrix

```{r}
cor_mat <- df |>
    dplyr::distinct(id, .keep_all = TRUE) |>
    select(-id, -site) |>
    cor()

round(cor_mat, 3)
```

```{r}
cor_mat |>
    as.data.frame() |>
    tibble::rownames_to_column("var1") |>
    tidyr::pivot_longer(-var1, names_to = "var2", values_to = "cor") |>
    ggplot(aes(x = var1, y = var2, fill = cor)) +
    geom_tile() +
    geom_text(aes(label = round(cor, 2)), size = 3) +
    scale_fill_gradient2(low = "#B2182B", mid = "white", high = "#2166AC",
                         limits = c(-1, 1)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = NULL, y = NULL, fill = "r")
```


## Is the distribution of sum(SNPs) normal?

```{r}
plot_df <-
df |>
  distinct(id, .keep_all = TRUE)
```

### Histogram

```{r}
plot_df |>
  ggplot() +
  aes(x = x_count) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 color = "black",
                 fill = "lightgray",
                 linewidth = 0.05) +
  geom_density() +
  theme_bw()
```

### ECDF

```{r}
mu <- mean(plot_df$x_count)
sigma <- sd(plot_df$x_count)

theoretical_df <- tibble(
    x_count = seq(min(plot_df$x_count), max(plot_df$x_count), length.out = 1000L),
    cdf = pnorm(x_count, mean = mu, sd = sigma)
)

ggplot(plot_df, aes(x = x_count)) +
    stat_ecdf(aes(color = "Empirical"), geom = "step") +
    geom_line(data = theoretical_df,
              aes(y = cdf, color = "Theoretical (Normal)"),
              linetype = "dashed") +
    scale_color_manual(values = c("Empirical" = "black",
                                  "Theoretical (Normal)" = "red")) +
    labs(x = "PRS (sum of SNPs)", y = "CDF", color = NULL) +
    theme_bw()
```

