---
title: "No Covariates: Imputation Model Comparison"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

## Setup

```{r}
#| label: setup
renv::load("/Users/max/r_package_development/bayes2stage")
library(ggplot2)
library(patchwork)
library(gt)
theme_set(cowplot::theme_cowplot(font_size = 12, font_family = "Source Sans Pro"))

SCRIPT_DIR <- normalizePath("..")
source("../lhs_config.R")
fs::dir_create("figures")
```

## Load Data

```{r}
#| label: load-data
df <- load_lhs_data()
N_SUBJECTS <- length(unique(df$id))
N_SAMPLED <- as.integer(SAMPLING_FRACTION * N_SUBJECTS)
```

The no-covariates specification uses `y ~ x + t + x:t + (1 + t | id)` with no additional covariates in either the outcome or imputation model. This isolates the effect of the imputation distribution choice without confounding from covariate adjustment.

## Load Pre-Computed Posterior Draws

```{r}
#| label: load-draws
result_files <- fs::dir_ls("cr_results", glob = "*.parquet")
cli::cli_alert_info("Found {length(result_files)} result files")

draws_df <- purrr::map(result_files, arrow::read_parquet) |>
    dplyr::bind_rows()

DESIGN_LABELS <- c(
    full          = "Full data",
    srs           = "SRS",
    srs_no_imp    = "SRS (no imp)",
    ods_intercept = "ODS (intercept)",
    ods_slope     = "ODS (slope)",
    ods_ellipse   = "ODS (ellipse)",
    bds_intercept = "BDS (intercept)",
    bds_slope     = "BDS (slope)",
    bds_ellipse   = "BDS (ellipse)"
)

DESIGN_ORDER <- c(
    "Full data", "SRS", "SRS (no imp)",
    "ODS (intercept)", "ODS (slope)", "ODS (ellipse)",
    "BDS (intercept)", "BDS (slope)", "BDS (ellipse)"
)

IMP_LABELS <- c(
    normal            = "Normal",
    beta_binomial     = "Beta-binomial",
    negative_binomial = "Negative-binomial"
)

IMP_ORDER <- c("Normal", "Beta-binomial", "Negative-binomial")

draws_df <- draws_df |>
    dplyr::mutate(
        design_label = factor(DESIGN_LABELS[design], levels = DESIGN_ORDER),
        imputation = factor(IMP_LABELS[imputation], levels = IMP_ORDER)
    )

cli::cli_alert_success(
    "Loaded {nrow(draws_df)} total draws across {length(unique(draws_df$imputation))} imputation models"
)
```

## Joint Credible Regions

```{r}
#| label: colors
IMP_COLORS <- c(
    "Normal"            = "#3574a7",
    "Beta-binomial"     = "#e8833a",
    "Negative-binomial" = "#6a9f58"
)
```

### Faceted by design, colored by imputation model

```{r}
#| label: fig-facet-comparison
#| fig-width: 16
#| fig-height: 10

p_facet <- draws_df |>
    ggplot(aes(x = beta_x, y = beta_x_t_interaction, color = imputation)) +
    stat_ellipse(level = 0.95, type = "norm", linewidth = 0.9) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey60") +
    geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
    scale_color_manual(values = IMP_COLORS) +
    facet_wrap(~ design_label, ncol = 4) +
    labs(
        title = "No covariates: joint 95% credible regions",
        subtitle = "No covariate adjustment | 25% sampling fraction",
        x = expression(beta[x] ~ "(PRS effect on FVC)"),
        y = expression(beta["x\u00d7t"] ~ "(PRS \u00d7 Time interaction)"),
        color = "Imputation model"
    )

p_facet

ggsave("figures/compare_imputation_faceted.png", p_facet,
    device = ragg::agg_png, bg = "white", dpi = 500,
    width = 16, height = 10, units = "in")
```

### Normal imputation: multiple credible levels

```{r}
#| label: fig-normal-levels
#| fig-width: 16
#| fig-height: 10

CR_LEVELS <- c(0.50, 0.80, 0.95)

p_levels <- draws_df |>
    dplyr::filter(imputation == "Normal") |>
    ggplot(aes(x = beta_x, y = beta_x_t_interaction)) +
    purrr::map(CR_LEVELS, \(lv) {
        stat_ellipse(
            level = lv, type = "norm", linewidth = 0.7,
            aes(color = factor(lv))
        )
    }) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey60") +
    geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
    scale_color_manual(
        values = c("0.5" = "#3574a7", "0.8" = "#e8833a", "0.95" = "#6a9f58"),
        labels = c("0.5" = "50%", "0.8" = "80%", "0.95" = "95%")
    ) +
    facet_wrap(~ design_label, ncol = 4) +
    labs(
        title = "Normal imputation: joint credible regions at multiple levels",
        subtitle = "No covariate adjustment | 25% sampling fraction",
        x = expression(beta[x] ~ "(PRS effect on FVC)"),
        y = expression(beta["x\u00d7t"] ~ "(PRS \u00d7 Time interaction)"),
        color = "Credible level"
    )

p_levels

ggsave("figures/compare_imputation_levels.png", p_levels,
    device = ragg::agg_png, bg = "white", dpi = 500,
    width = 16, height = 10, units = "in")
```

### ODS vs BDS overlay

```{r}
#| label: fig-overlay-design
#| fig-width: 14
#| fig-height: 6

TARGET_COLORS <- c(
    "Intercept" = "#3574a7",
    "Slope"     = "#e8833a",
    "Ellipse"   = "#6a9f58"
)

draws_normal <- draws_df |>
    dplyr::filter(imputation == "Normal")

draws_full <- draws_normal |>
    dplyr::filter(design == "full")

draws_ods_bds <- draws_normal |>
    dplyr::filter(grepl("^(ods|bds)", design)) |>
    dplyr::mutate(
        design_family = ifelse(grepl("^ods", design), "ODS", "BDS"),
        design_family = factor(design_family, levels = c("ODS", "BDS")),
        target = dplyr::recode_values(
            design,
            c("ods_intercept", "bds_intercept") ~ "Intercept",
            c("ods_slope", "bds_slope") ~ "Slope",
            c("ods_ellipse", "bds_ellipse") ~ "Ellipse"
        ),
        target = factor(target, levels = c("Intercept", "Slope", "Ellipse"))
    )

p_overlay <- draws_ods_bds |>
    ggplot(aes(x = beta_x, y = beta_x_t_interaction, color = target)) +
    stat_ellipse(
        data = draws_full |> tidyr::crossing(
            design_family = factor(c("ODS", "BDS"), levels = c("ODS", "BDS"))
        ),
        aes(x = beta_x, y = beta_x_t_interaction),
        level = 0.95, type = "norm", linewidth = 0.7,
        linetype = "dashed", color = "grey40", inherit.aes = FALSE
    ) +
    stat_ellipse(level = 0.95, type = "norm", linewidth = 0.9) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey60") +
    geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
    scale_color_manual(values = TARGET_COLORS) +
    facet_wrap(~ design_family) +
    labs(
        title = "Sampling target comparison: joint 95% credible regions",
        subtitle = "Normal imputation | No covariate adjustment | 25% sampling fraction | Dashed = full data",
        x = expression(beta[x] ~ "(PRS effect on FVC)"),
        y = expression(beta["x\u00d7t"] ~ "(PRS \u00d7 Time interaction)"),
        color = "Target"
    )

p_overlay

ggsave("figures/compare_imputation_overlay.png", p_overlay,
    device = ragg::agg_png, bg = "white", dpi = 500,
    width = 14, height = 6, units = "in")
```

### Full data only

```{r}
#| label: fig-full-comparison
#| fig-width: 8
#| fig-height: 6

p_full <- draws_df |>
    dplyr::filter(design == "full") |>
    ggplot(aes(x = beta_x, y = beta_x_t_interaction, color = imputation)) +
    stat_ellipse(level = 0.95, type = "norm", linewidth = 0.9) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey60") +
    geom_vline(xintercept = 0, linetype = "dotted", color = "grey60") +
    scale_color_manual(values = IMP_COLORS) +
    labs(
        title = "Full data: no covariates, imputation model comparison",
        x = expression(beta[x] ~ "(PRS effect on FVC)"),
        y = expression(beta["x\u00d7t"] ~ "(PRS \u00d7 Time interaction)"),
        color = "Imputation model"
    )

p_full

ggsave("figures/compare_imputation_full.png", p_full,
    device = ragg::agg_png, bg = "white", dpi = 500,
    width = 8, height = 6, units = "in")
```

## Marginal Comparisons

```{r}
#| label: fig-marginal
#| fig-width: 16
#| fig-height: 10

p_bx <- draws_df |>
    ggplot(aes(x = beta_x, fill = imputation)) +
    geom_density(alpha = 0.35, color = NA) +
    scale_fill_manual(values = IMP_COLORS) +
    facet_wrap(~ design_label, ncol = 4, scales = "free_y") +
    labs(
        title = expression(bold("Marginal posterior:") ~ beta[x]),
        x = expression(beta[x]),
        y = "Density",
        fill = "Imputation"
    )

p_bxt <- draws_df |>
    ggplot(aes(x = beta_x_t_interaction, fill = imputation)) +
    geom_density(alpha = 0.35, color = NA) +
    scale_fill_manual(values = IMP_COLORS) +
    facet_wrap(~ design_label, ncol = 4, scales = "free_y") +
    labs(
        title = expression(bold("Marginal posterior:") ~ beta["x\u00d7t"]),
        x = expression(beta["x\u00d7t"]),
        y = "Density",
        fill = "Imputation"
    )

p_marginal <- p_bx / p_bxt + plot_layout(guides = "collect") &
    theme(legend.position = "bottom")

p_marginal

ggsave("figures/compare_imputation_marginal.png", p_marginal,
    device = ragg::agg_png, bg = "white", dpi = 500,
    width = 16, height = 10, units = "in")
```

## Forest Plots

```{r}
#| label: summarize-draws
prob_df <- draws_df |>
    dplyr::summarize(
        median_bx  = median(beta_x),
        median_bxt = median(beta_x_t_interaction),
        q2_5_bx    = quantile(beta_x, 0.025),
        q97_5_bx   = quantile(beta_x, 0.975),
        q2_5_bxt   = quantile(beta_x_t_interaction, 0.025),
        q97_5_bxt  = quantile(beta_x_t_interaction, 0.975),
        p_bx_pos   = mean(beta_x > 0),
        p_bx_small_01  = mean(abs(beta_x) < 0.01),
        p_bx_small_005 = mean(abs(beta_x) < 0.005),
        p_bxt_neg  = mean(beta_x_t_interaction < 0),
        p_joint    = mean(abs(beta_x) < 0.005 & beta_x_t_interaction < 0),
        .by = c(design_label, imputation)
    )

make_forest_data <- function(prob_data, est_col, lo_col, hi_col) {
    designs <- levels(prob_data$design_label)
    imps <- levels(prob_data$imputation)

    full_ref <- dplyr::filter(prob_data,
        design_label == "Full data", imputation == "Normal")
    full_se <- (full_ref[[hi_col]] - full_ref[[lo_col]]) / (2 * qnorm(0.975))

    rows <- list()
    for (des in designs) {
        rows <- c(rows, list(
            data.frame(label = des, est = NA_real_, lo = NA_real_,
                       hi = NA_real_, se = NA_real_, re = NA_real_)
        ))
        available_imps <- dplyr::filter(prob_data, design_label == des)$imputation
        for (imp in intersect(imps, available_imps)) {
            r <- dplyr::filter(prob_data,
                design_label == des, imputation == imp)
            this_se <- (r[[hi_col]] - r[[lo_col]]) / (2 * qnorm(0.975))
            rows <- c(rows, list(data.frame(
                label = imp,
                est = r[[est_col]],
                lo  = r[[lo_col]],
                hi  = r[[hi_col]],
                se  = this_se,
                re  = (full_se / this_se)^2
            )))
        }
        rows <- c(rows, list(
            data.frame(label = "", est = NA_real_, lo = NA_real_,
                       hi = NA_real_, se = NA_real_, re = NA_real_)
        ))
    }
    dplyr::bind_rows(rows)
}
```

### $\beta_x$ (PRS effect on FVC)

```{r}
#| label: fig-forest-bx
#| fig-width: 12
#| fig-height: 12

fd_bx <- make_forest_data(prob_df, "median_bx", "q2_5_bx", "q97_5_bx")

full_est <- dplyr::filter(prob_df, design_label == "Full data",
                           imputation == "Normal")$median_bx

p_bx <- bayes2stage::plot_forest(
    fd_bx, ref_line = full_est,
    xlab = expression(beta[x]),
    ci_format = "%.4f (%.4f, %.4f)", se_format = "%.4f",
    extra_col = list(col = "re", header = "RE", format = "%4.2f"),
    font_family = ""
) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
    ggtitle(
        expression(beta[x] ~ ": No covariates model"),
        subtitle = "Posterior median (95% CrI) | 25% sampling fraction"
    ) +
    theme(
        plot.title = element_text(size = 13, face = "bold"),
        plot.title.position = "plot",
        plot.subtitle = element_text(size = 10)
    )

p_bx

ggsave("figures/compare_forest_bx.png", p_bx,
    device = ragg::agg_png, bg = "white", dpi = 500,
    width = 12, height = 12, units = "in")
```

### $\beta_{x \times t}$ (PRS $\times$ Time interaction)

```{r}
#| label: fig-forest-bxt
#| fig-width: 12
#| fig-height: 12

fd_bxt <- make_forest_data(prob_df, "median_bxt", "q2_5_bxt", "q97_5_bxt")

full_est_bxt <- dplyr::filter(prob_df, design_label == "Full data",
                                imputation == "Normal")$median_bxt

p_bxt <- bayes2stage::plot_forest(
    fd_bxt, ref_line = full_est_bxt,
    xlab = expression(beta["x\u00d7t"]),
    ci_format = "%.5f (%.5f, %.5f)", se_format = "%.5f",
    extra_col = list(col = "re", header = "RE", format = "%4.2f"),
    font_family = ""
) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
    ggtitle(
        expression(beta["x\u00d7t"] ~ ": No covariates model"),
        subtitle = "Posterior median (95% CrI) | 25% sampling fraction"
    ) +
    theme(
        plot.title = element_text(size = 13, face = "bold"),
        plot.title.position = "plot",
        plot.subtitle = element_text(size = 10)
    )

p_bxt

ggsave("figures/compare_forest_bxt.png", p_bxt,
    device = ragg::agg_png, bg = "white", dpi = 500,
    width = 12, height = 12, units = "in")
```

## Probability Statement Comparison

```{r}
#| label: tbl-comparison
tbl_prob <- prob_df |>
    dplyr::mutate(
        design_family = dplyr::case_when(
            design_label == "Full data" ~ "Full",
            design_label %in% c("SRS", "SRS (no imp)") ~ "SRS",
            grepl("^ODS", design_label) ~ "ODS",
            grepl("^BDS", design_label) ~ "BDS"
        ),
        target = dplyr::recode_values(
            as.character(design_label),
            c("Full data", "SRS", "SRS (no imp)") ~ "\u2014",
            c("ODS (intercept)", "BDS (intercept)") ~ "Intercept",
            c("ODS (slope)", "BDS (slope)") ~ "Slope",
            c("ODS (ellipse)", "BDS (ellipse)") ~ "Ellipse"
        ),
        design_family = factor(design_family, levels = c("Full", "SRS", "ODS", "BDS")),
        target = factor(target, levels = c("\u2014", "Intercept", "Slope", "Ellipse"))
    ) |>
    dplyr::arrange(design_family, target, imputation) |>
    dplyr::select(design_family, target, imputation,
        p_bx_pos, p_bx_small_01, p_bx_small_005, p_bxt_neg, p_joint) |>
    gt(groupname_col = "design_family") |>
    fmt_percent(columns = starts_with("p_"), decimals = 1) |>
    cols_label(
        target         = "Target",
        imputation     = "Imputation",
        p_bx_pos       = gt::md("P(&beta;~x~ > 0)"),
        p_bx_small_01  = gt::md("P(|&beta;~x~| < 0.01)"),
        p_bx_small_005 = gt::md("P(|&beta;~x~| < 0.005)"),
        p_bxt_neg      = gt::md("P(&beta;~x&times;t~ < 0)"),
        p_joint        = gt::md("P(|&beta;~x~| < 0.005 &amp; &beta;~x&times;t~ < 0)")
    )

tbl_prob

gtsave(tbl_prob, "figures/compare_tbl_probabilities.png", zoom = 3)
```

## Diagnostics

```{r}
#| label: tbl-diagnostics
diag_df <- draws_df |>
    dplyr::distinct(design_label, imputation, divergent, max_treedepth, ebfmi_min)

tbl_diag <- diag_df |>
    gt(groupname_col = "design_label") |>
    cols_label(
        imputation    = "Imputation",
        divergent     = "Divergent",
        max_treedepth = "Max treedepth",
        ebfmi_min     = "Min E-BFMI"
    ) |>
    fmt_number(columns = ebfmi_min, decimals = 2)

tbl_diag

gtsave(tbl_diag, "figures/compare_tbl_diagnostics.png", zoom = 3)
```
